### 2014-04-29 text/Chapter03.tex
Tweak style and finish Chapter 3

\subsection{Tainting and provenance}
Tainting is a mechanism where variables coming from potentially untrusted sources are marked
(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
values are annotated with more detailed information about their source.

Statically typed systems that based on tainting have been use to prevent cross-site scripting
attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
\ident{msg}~+~\ident{name}
\end{array}
\end{equation*}
%
In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
untainted string. Otherwise, the attacker could specify values such as \str{"1; DROP TABLE Users"}. 
The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
(i.e.~to construct an SQL query).

In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
flag (for tainting) or more complex data (for provenance). In static checking, the information
need to be associated with the variables in the variable context. We use tainting as a motivating
example for \emph{structural} coeffects in Section~X.

@@

\subsection{Security and core dependency calculus}

The checking of tainting is a special case of checking of the \emph{non-interference} property 
in \emph{secure information flow}. Here, the aim is to guarantee that sensitive information (such
as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
more recent work. The checking of information flows has been also integrated (as a single-purpose
extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
using a monadic library.

@@

An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
example which may assign a new value to $z$:
%
\begin{equation*}
\kvd{if}~x>0~\kvd{then}~z := y
\end{equation*}
%
If the value of $y$ is high-secure, then $z$ becomes high-secure after the assignment
(this is an \emph{explicit} flow). However, if $x$ is high-secure, then the value of
$z$ becomes high-secure, regardless of the security level of $y$, because the fact whether an 
assignment is performed or not performed leaks information in its own (this is an 
\emph{implicit} flow).

@@

@@

In the systems discussed so far, the context provides additional data (resources, implicit 
parameters, historical values) or meta-data (security, provenance). However, it is impossible to
write a function that modifies the context. We use the term \emph{passive} context for such 
applications.

However, there is a number of systems where the context may be changed -- not just be evaluating
certain code block in a different scope (e.g. by wrapping it in $\ident{prev}$ in data-flow), but
also by calling a function that, for example, acquires new capabilities. While this thesis focuses
on systems with passive context, we quickly look at the most important examples of the 
\emph{active} variant.

@@

\paragraph{Calculus of capabilities}
\label{sec:applications-active-ccc}

Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
a sound system with region-based memory management for low-level code that can be easily 
compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
who developed an \emph{effect system} (discussed in Section~\ref{sec:path-sem-effects}) that uses
lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.

In the work of Tofte and Talpin, the context is \emph{passive}. They extend a simple functional language
with the \kvd{letrgn} construct that defines a new memory region, evaluates an expression (possibly)
using memory in that region and then deallocates the memory of the region:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
\quad\kvd{letrgn}~\rho~\kvd{in}\\
\quad\kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~!\ident{x}
\end{array}
\end{equation*}
%
The memory region $\rho$ is a part of the context, but only in the scope of the body of 
\kvd{letrgn}. It is only available to the last line which allocates a memory cell in the region
and reads it (before the region is deallocated). There is no way to allocate a region inside a 
function and pass it back to the caller.

Calculus of capabilities differs in two ways. First, it allows explicit allocation and deallocation
of memory regions (and so region lifetimes do not follow strict LIFO ordering). Second, it
uses continuation-passing style. We ignore the latter aspect and so the following example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
\quad \kvd{letrgn}~\rho~\kvd{in}\\
\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~\ident{x}
\end{array}
\end{equation*}
%
The example is almost identical to the previous one, except that it does not return the value
of reference \ident{x}. Instead, it returns the reference, which is located in a newly allocated
region. Together with the value, the function returns a \emph{capability} to access the region
$\rho$.

This is where systems with active context differ. To type check such programs, we do not only need
to know what context is required to call \ident{calculate}. We also need to know what effects it
has on the context when it evaluates and the current context meeds to be appropriately adjusted
after a function call. We briefly consider this problem in Section~X. % future work

@@

\paragraph{Software updating}
Dynamic software updating (DSU) \cite{app-dsu-programs,app-dsu} is the ability to update programs at
runtime without stopping them. The Proteus system developed by Stoyle et al. \cite{app-dsu-mutatis} 
investigates what language support is needed to enable safe dynamic software updating in C-like 
languages. The system is based on the idea of capabilities.

@@

\paragraph{Thesis perspective}

As demonstrated in this section, there is a huge number of systems and applications that exhibit
a form of context-dependence. The range includes different static analyses (liveness, provenance), 
well-known programming language features (implicit parameters and type classes) as well as features
not widely available (e.g. for distributed programming).

It is impossible to cover all of these topics in a single coherent thesis and so we focus on 
two key aspects:

\begin{compactitem}
\item \textbf{Flat vs. structural.} We look at both flat coeffects (single value for entire context) and 
  structural coeffects (single value per variable). We use liveness, implicit parameters and 
  data-flow to introduce flat coeffects (Section~X) and liveness, refined data-flow and tainting
  to talk about structural coeffects (Section~Y).

\item \textbf{Analysis vs. restriction.} Some of the discussed examples can be viewed as static
  analyses that obtain some information about programs (i.e.~the number of required past values
  in data-flow). Other examples provide type system that rules out certain invalid programs 
  (e.g.~safe locking). We cover this topic when discussing \emph{partial coeffects} in Section~Z.

\item \textbf{May vs. must analysis.} When discussing liveness, we observed that we can obtain 
  two different analyses depending on how conditionals are treated. We discuss this topic in 
  Section~X. % TBD - could be a chapter
\end{compactitem}

Although we also looked at examples of \emph{active} contextual computations (where developers can
write functions that modify the context), we do not consider these applications, to keep the 
material presented in this thesis focused. We briefly discuss them as future work in Section~X.

@@

TODO

% \section{Missing}
% ~
% 
% indexed/layered/etc. monads, productors or whatever

>>>>

\subsection{Security, tainting and provenance}
Tainting is a mechanism where variables coming from potentially untrusted sources are marked
(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
values are annotated with more detailed information about their source.

Statically typed systems that based on tainting have been use to prevent cross-site scripting
attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{name} = \ident{query}(\str{SELECT Name WHERE Id = \%1}, \ident{id})\\[-0.25em]
\ident{msg}~+~\ident{name}
\end{array}
\end{equation*}
%
In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
untainted string. Otherwise, the attacker could specify values such as \str{1; DROP TABLE Users}. 
The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
(i.e.~to construct an SQL query).

In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
flag (for tainting) or more complex data (for provenance). In static checking, the information
need to be associated with the variables in the variable context.

@@

\paragraph{Core dependency calculus.}
The checking of tainting is a special case of checking of the \emph{non-interference} property 
in \emph{secure information flow}. There, the aim is to guarantee that sensitive information (such
as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
more recent work. The checking of information flows has been also integrated (as a single-purpose
extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
using a monadic library.

@@

\paragraph{Implicit flows.}
An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
example which returns either $y$ or zero, depending on the value of $x$:
%
\begin{equation*}
\kvd{let}~z = \kvd{if}~x>0~\kvd{then}~y~\kvd{else}~0
\end{equation*}
%
If the value of $y$ is high-secure, then $z$ becomes high-secure after the assignment
(this is an \emph{explicit} flow). However, if $x$ is high-secure, then the value of
$z$ becomes high-secure, regardless of the security level of $y$, because the fact whether an 
assignment is performed or not performed leaks information in its own (this is an 
\emph{implicit} flow).

@@

\paragraph{Coeffect systems.}
The work outlined in this section is another area where coeffect systems could be applied.
We do not develop coeffect systems for tracking of tainting, security and provenance in details,
but briefly mention some examples in the upcoming chapters.

The systems work in the same way as the examples discussed already. For example, consider the
tainting example with the \ident{query} function calling an SQL database. To capture such 
tainting, we annotate variables with $\cclrd{\ident{T}}$ for \emph{tainted} and with 
$\cclrd{\ident{U}}$ for \emph{untainted}. Simply accessing variable does not introduce taint, 
but using a variable in certain contexts -- such as in arguments of \ident{query} -- does introduce
a taint. This is captured using the standard application rule (\emph{app}):
%
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma}{\aclrd{r}} \vdash \ident{query} : \ident{string} \xrightarrow{\cclrd{\ident{T}}} \ident{Table} \qquad
    \coctx{\ident{id}:\ident{string}}{ \alift{\cclrd{\ident{U}}}} \vdash \ident{id} : \ident{string} }
  { \coctx{\Gamma, \ident{id}:\ident{string}}{ \aclrd{r} \cons \alift {\cclrd{\ident{T}}} \vdash \ident{query}(\str{...}, \ident{id}) : \ident{Table} } }
\end{equation*}
%
The derivation assumes that \ident{query} is a standard function that requires the parameters
to be tainted (it does not have to be a built-in language construct). The argument is a 
variable and so it is not tainted in the assumptions.

In the conclusion, we need to derive an annotation for the variable \ident{id}. To do this, we
combine \cclrd{\ident{T}} (from the function) and \cclrd{\ident{U}} (from the argument). In case
of tainting, the variable is tainted whenever it is already tainted \emph{or} the function marks
it as tainted. For different kinds of annotations, the composition would work differently -- for
example, for provenance, we could union the \emph{set} of possible data sources, or even combine 
\emph{probability distributions} modelling the influence of different sources on the value.
However, expanding such ideas is beyond the scope of this thesis.

@@

In both flat and structural systems discussed so far, the context provides additional data (resources, 
implicit parameters, historical values) or meta-data (security, provenance). However, \emph{within} 
the language, it is impossible to write a function that modifies the context. We use the term 
\emph{passive} context for such applications.

There is a number of systems that also capture contextual properties, but that make it possible to 
\emph{change} the context -- not just be evaluating certain code block in a different scope (e.g. by 
wrapping it in $\ident{prev}$ in data-flow), but also by calling a function that, for example, acquires 
new capabilities and returns those to the caller. While this thesis focuses on systems with passive 
contexts, we briefly consider the most important examples of the \emph{active} variant.

@@

\paragraph{Calculus of capabilities.}
Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
a sound system with region-based memory management for low-level code that can be easily 
compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
who developed an effect system (as discussed in Section~\ref{sec:path-sem-effects}) that uses
lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.

In the work of Tofte and Talpin, the context is \emph{passive}. They extend a simple functional language
with the \kvd{letrgn} construct that defines a new memory region, evaluates an expression (possibly)
using memory in that region and then deallocates the memory of the region:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\[-0.25em]
\quad \kvd{letrgn}~\rho~\kvd{in}\\[-0.25em]
\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho ~\ident{input}~\kvd{in}\\[-0.25em]
\quad \ident{x} :=\, !\ident{x} + 1;~ !\ident{x}
\end{array}
\end{equation*}
%
The memory region $\rho$ is a part of the context, but only in the scope of the body of 
\kvd{letrgn}. It is only available to the last two lines which allocate a memory cell in the region,
increment a value in the region and then read it. The region is de-allocated when the execution 
leaves its lexical scope -- there is no way to allocate a region inside a function and pass it back 
to the caller.

Calculus of capabilities differs in two ways. First, it allows explicit allocation and deallocation
of memory regions (and so region lifetimes do not follow strict LIFO ordering). Second, it
uses continuation-passing style. We ignore the latter aspect. The following example is almost 
identical as the previous one:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\[-0.25em]
\quad \kvd{letrgn}~\rho~\kvd{in}\\[-0.25em]
\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho ~\ident{input}~\kvd{in}\\[-0.25em]
\quad \ident{x} :=\, !\ident{x} + 1;~ \ident{x}
\end{array}
\end{equation*}
%
The difference is that the example does not return the \emph{value} of a reference using 
$!\ident{x}$, but returns the reference $\ident{x}$ itself. The reference is allocated in a newly 
created region $\rho$. Together with the value, the function returns a \emph{capability} to access 
the region $\rho$.

This is where systems with active context differ. To type check such programs, we do not only need
to know what context is required to call \ident{calculate}. We also need to know what effects it
has on the context when it evaluates and the current context meeds to be updated after a function 
call.

\paragraph{Active contexts.}
In a systems with passive contexts, we only need an annotation that specifies the required 
context. In semantics, this is reflected by having some structure (data type) $\C$ over the
\emph{input} of the function. Without giving any details, the semantics generally has the 
following structure:
%
\begin{equation*}
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} = \C^{\cclrd{r}} \tau_1 \rightarrow \tau_2
\end{equation*}
%
Systems with active contexts require two annotations -- one that specifies the context required
before the call is performed and one that specifies how the context changes after the call (this
could be either a \emph{new} context or \emph{update} to the original context). Thus the structure
of the semantics would look as follows:
%
\begin{equation*}
\sem{\tau_1 \xrightarrow{\cclrd{r}, \cclrd{s}} \tau_2} = \C^{\cclrd{r}} \tau_1 \rightarrow \M^{\cclrd{s}} \tau_2
\end{equation*}
%
In case of Calculus of Capabilities, both of the structures could be the same and they could
carry a set of available memory regions. In this thesis, we focus only on passive contexts. 
However, we briefly consider the problem of active contexts in the Section~X of the future work 
chapter.

@@

\paragraph{Software updating.}
Another example of a system that uses contextual information actively is dynamic software updating 
(DSU) \cite{app-dsu-programs,app-dsu}. The DSU systems have the ability to update programs at
runtime without stopping them. For example, Proteus developed by Stoyle et al. \cite{app-dsu-mutatis} 
investigates what language support is needed to enable safe dynamic software updating in C-like 
languages. The system is based on the idea of capabilities.

@@

@@

This chapter served two purposes. The first aim was to present existing work on programming 
languages and systems that include some notion of \emph{context}. Because there was no well-known
abstraction capturing contextual properties, the languages use a wide range of formalisms -- including
from principled approaches based on comonads, ad-hoc type system extensions and static analyses as
well as approaches based on monads. We looked at a number of applications inclding Haskell's implicit 
parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of 
security properties.

The second aim of this chapter was to re-formulate the existing work in a more uniform style and thus
reveal that all \emph{context-dependent} languages share a common structure. In the upcoming three 
chapters, we identify the common structure more precisely and develop three calculi to capture it. We 
will then be able to re-create many of the examples discussed in this chapter by instantiating our 
unified calculi.

This chapter was divided into two major sections. First, we looked at \emph{flat} systems, which
track whole-context properties. Next, we look a \emph{structural} systems, which track per-variable
properties. Both of the variants are useful and important -- for example, implicit parameters can
only be expressed as \emph{flat} system, but liveness analysis is only useful as \emph{structural}.
For this reason, we explore both of these variants in this thesis (Chapter~\ref{ch:flat} and 
Chapter~\ref{ch:structural}, respectively). We can, however, unify the two variants into a single
system discussed in Chapter~\ref{ch:unified}.
